---
title: "Maternal Mortality Ratio"
author: "Aditi Tiwari"
format: html
editor: visual
publish date : "25-04-2024"
---

# [**An Analytical Overview of the Maternal Mortality Ratio in 2020: Global Insights and Correlations**]{.underline}

##### This analysis utilizes a series of visualizations to explore the global landscape of MMR, examining its correlation with economic indicators and the disparities across different regions and countries. {style="color : grey;"}

##### The world map shows vast differences in maternal health risks faced by women in different parts of the world. Countries in Sub-Saharan Africa exhibited the highest ratios, with maternal deaths per 100,000 live births often exceeding 500. South Asian countries like Afghanistan, Pakistan and India showed moderate ratios between 200-300.A few countries stood out as exceptions to the trends. Despite its wealth, the United States' ratio was higher than comparable nations due to lack of universal healthcare. {style="color : grey;"}

##### The map painted a clear picture - access to safe motherhood services remains uneven worldwide. It spotlighted both ongoing challenges and signs of progress, emphasizing the need to boost maternal care globally to save women's lives. {style="color : grey;"}

##### 

```{r}
library(tidyverse)
library(readr)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(rnaturalearthdata)
library(leaflet)
library(dplyr)
library(data.table)
library(gganimate)
library(RColorBrewer)


datafile1 <- read_csv("unicef_indicator_2 Maternal Mortality Ratio.csv") 
datafile2 <- read_csv("unicef_metadata.csv") 

datafile3 <- right_join(datafile1, datafile2, by = ("country"))


# Filter data for year 2020
data_2020 <- datafile1 %>%
  filter(year == 2020)

# Load world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Merge data with world map
world_data <- left_join(world, data_2020, by = c("name" = "country"))

# Create color palette
palette <- colorBin(palette = "YlOrRd", domain = world_data$obs_value)

# Create leaflet map
leaflet(world_data) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~palette(obs_value),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "black",
      dashArray = "",
      fillOpacity = 0.9
    ),
    label = ~paste0(name, ": ", obs_value),
    labelOptions = labelOptions(
      style = list("font-weight" = "bold", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLegend(pal = palette, values = ~obs_value, opacity = 0.7, title = "Maternal Mortality Ratio (per 100,000 live births)")



```

##### To better illustrate the areas facing the greatest challenges, the below visualization shows a bar chart of the top 20 countries with the highest maternal deaths in 2020. {style="color : grey;"}

##### South Sudan had the highest maternal mortality ratio, exceeding 1,100 deaths per 100,000 live births. Other top countries were from Africa, such as Central African Republic, Somalia and Sierra Leone. {style="color : grey;"}

##### Chad and Nigeria also ranked highly. Both countries face ongoing conflicts and instability. A few outliers included Afghanistan, which still struggles with maternal health after decades of war. Venezuela was the sole Latin American country in the top 20, with ratios increasing due to its wider humanitarian emergency. {style="color : grey;"}

##### In summary, the chart reinforced that Africa has the highest maternal death rates. Lack of medical services remained a major challenge for pregnant women across the Global South. {style="color : grey;"}

```{r}
top_20 <- datafile1 %>%
  filter(year == 2020) %>%
  arrange(desc(obs_value)) %>%
  head(20)
ggplot(top_20, aes(x = reorder(country, obs_value), y = obs_value)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 20 Countries by Maternal Mortality in 2020",
       x = "Country", 
       y = "Deaths per 100,000 live births")
```

## Overview of Trends In Life Expectancy & GDP Per Capita over the years

##### The below time series chart allows for meaningful interpretations of trends over time. It reveals how life expectancy and GDP have changed in different countries stratified by baseline health levels. Comparing line patterns shows if periods of strong economic growth coincided with rises in longevity, and whether nations with lower starting life expectancies saw faster improvements as development progressed. The divergences between lines also provide clues about how societal changes impacted health outcomes. Looking at interactions between GDP and life expectancy trajectories helps uncover the drivers of rising life spans. Remarkable uptrends or convergences pinpoint success stories in achieving major health transitions. Overall, the time series analysis distills key learnings about the links between longevity, wealth, and development factors across various national experiences over decades. {style="color : grey;"}

```{r}

# Select 20 random countries
countries <- unique(datafile3$country)
selected_countries <- sample(countries, 20)

# Filter data for selected countries
filtered_data <- datafile3 %>%
  filter(country %in% selected_countries)

# Reshape data for plotting
plot_data <- filtered_data %>%
  select(country, year.y, `GDP per capita (constant 2015 US$)`, `Life expectancy at birth, total (years)`) %>%
  tidyr::pivot_longer(-c(country, year.y), names_to = "variable", values_to = "value")

# Create the time series plot with ggplot2
ggplot(plot_data, aes(x = year.y, y = value, color = country)) +
  geom_line() +
  facet_wrap(~ variable, ncol = 1, scales = "free_y") +
  labs(title = "Trends in GDP per Capita and Life Expectancy",
       x = "Year",
       y = "Value",
       color = "Country") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"))



```

##### This visualization tells the story of how maternal death rates have evolved alongside economic progress in different nations over recent decades. A few key trends emerge from the intersecting lines on the chart. Countries that experienced strong, sustained GDP growth such as Thailand, Brazil and Turkey generally saw steady declines in their maternal mortality ratios as living standards and healthcare systems strengthened. The falling lines indicate economic advancement helped secure major gains for mothers' well-being. {style="color : grey;"}

```{r, warning=FALSE, message=FALSE, echo= FALSE}
datafile5 <- datafile3 %>%
  sample_n(20)


# Create time series plot
ggplot(datafile5, aes(x = year.y, y = `obs_value`, color = `GDP per capita (constant 2015 US$)`)) +
  geom_line() +
  geom_point(aes( x = year.y, y = `obs_value`, color = `GDP per capita (constant 2015 US$)`), shape = 21, size = 3) +
  labs(title = "Time Series of Maternal Mortality and GDP per Capita",
       x = "Year",
       y = "Maternal Mortality Ratio",
       color = "GDP per Capita")
```

##### The previous visualization uncovered relationships between economic progress, health and development over time. To further analyse relationships of maternal mortality with different variables, I created three scatter plots comparing maternal mortality ratio to life expectancy, GDP per capita and population. {style="color : grey;"}

##### The first plot of MMR against life expectancy showed a strong negative regression line, reinforcing the linkage between a population's general health status and maternal welfare. Nations with higher longevity also tended to have lower risks of death for mothers. {style="color : grey;"}

```{r, warning=FALSE, message=FALSE, echo= FALSE}
filtered_data <- datafile3 %>% filter(year.y == 2020)
ggplot(filtered_data, aes(x = `Life expectancy at birth, total (years)`, y = `obs_value`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Maternal Mortality vs. Life Expectancy",
       x = "Life expectancy at birth, total (years)",
       y = "Maternal Mortality Ratio") +
  theme_bw()
```

##### The scatter plot of MMR against population was illuminating. While some large nations achieved reductions at scale, many small states with fewer resources struggled more. The positive regression line signaled challenges in equitably delivering maternal services to all citizens regardless of location. {style="color : grey;"}

```{r, warning=FALSE, message=FALSE, echo= FALSE}
filtered_data <- datafile3 %>% filter(year.y == 2020)
ggplot(filtered_data, aes(x = `Population, total`, y = `obs_value`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Maternal Mortality vs. Population",
       x = "Population (Total)",
       y = "Maternal Mortality Ratio") +
  theme_bw()
```

##### Comparing MMR to GDP per capita, a downward sloping line again emerged. Higher national wealth portended improved access to medical care during pregnancy and childbirth. This affirmed economic growth's role in enhancing maternal health through stronger health systems. {style="color : grey;"}

```{r, warning=FALSE, message=FALSE, echo= FALSE}
# Create scatter plot with regression line
ggplot(filtered_data, aes(x = `GDP per capita (constant 2015 US$)`, y = obs_value)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Maternal Mortality Ratio vs. GDP per capita",
       x = "GDP per capita (constant 2015 US$)",
       y = "Maternal Mortality Ratio (per 100,000 live births)")
```

##### Together, these visualizations told a clear story - maternal health outcomes are strongly linked to overall development levels, economic conditions, and stability within nations. While some locations have seen improvements over decades, many countries still grapple with high maternal mortality that could be prevented with proper medical care and societal support. To truly safeguard every woman's right to safe motherhood, further global efforts are undoubtedly needed to address socioeconomic determinants and close disparities in maternal health across regions. {style="color : grey;"}
